---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Rzarrs

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rzarrs/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rzarrs/actions/workflows/R-CMD-check.yaml)
[![Rzarrs status badge](https://sounkou-bioinfo.r-universe.dev/Rzarrs/badges/version)](https://sounkou-bioinfo.r-universe.dev/Rzarrs)
<!-- badges: end -->

R bindings to the [`zarrs`](https://github.com/zarrs/zarrs) Rust library for
reading Zarr V3 (and compatible Zarr V2) stores. The current package is
reader-first: local filesystems, HTTP/HTTPS, S3, and local `.zarr.zip` VCF
archives are supported by default. GCS and Azure Blob can be enabled when
installing from source.

## Overview

Rzarrs exposes four savvy-backed reference objects:

| Object | Purpose |
|---|---|
| `ZarrStore` | Local filesystem store |
| `ZarrObjectStore` | HTTP/HTTPS and S3 by default; GCS/Azure Blob when enabled at source-install time |
| `ZarrGroup` | A group node within a store (attributes + child listing) |
| `ZarrArray` | A single array within any store |

Zarr dtypes are mapped to R types automatically:

| Zarr dtype | R type | Notes |
|---|---|---|
| `float32` / `float64` | `double` | NaN, Inf, -Inf preserved as-is |
| `int8` / `int16` / `int32` | `integer` | `i32::MIN` → `NA_integer_` |
| `int64` | `double` | exact to 2^53 |
| `uint8` / `uint16` | `integer` | always fits |
| `uint32` / `uint64` | `double` | |
| `bool` | `logical` | |
| `string` | `character` | variable-length UTF-8 via `vlen-bytes` codec |

Indices are **1-based and inclusive** on both ends — the same convention as
all other R array operations.

SIMD-accelerated codec paths (gzip, zstd, crc32c) are selected
automatically at runtime by the underlying Rust dependency crates.

## Installation

```r
install.packages(
  "Rzarrs",
  repos = c(
    "https://sounkou-bioinfo.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

A Rust toolchain (cargo + rustc >= 1.82) and GNU Make are required at install time.

Default Rust features are `aws` and `zip`: HTTP/HTTPS, S3, and local
`.zarr.zip` VCF Zarr archives work out of the box. Source installs can enable
more providers with configure arguments:

```r
# Enable GCS in addition to defaults
install.packages(
  "Rzarrs",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  configure.args = "--enable-gcp"
)

# Enable all cloud providers: AWS, GCS, and Azure Blob
install.packages(
  "Rzarrs",
  repos = c("https://sounkou-bioinfo.r-universe.dev", "https://cloud.r-project.org"),
  configure.args = "--enable-all-cloud"
)

# Exact Cargo feature control; comma or space separated features are accepted
install.packages(
  "Rzarrs",
  type = "source",
  configure.args = "--without-default-rust-features --with-rust-features=aws,gcp,azure,zip"
)
```

Equivalent environment-variable control is also supported:

```sh
SAVVY_FEATURES="aws gcp azure zip" R CMD INSTALL Rzarrs_0.1.0.tar.gz
```

## Local store

```{r local}
library(Rzarrs)

# The package ships a tiny bundled fixture for illustration
path <- system.file("testdata", "int32.zarr", package = "Rzarrs")

store <- ZarrStore$open(path)
store$path()

arr <- ZarrArray$open(store, "/")
arr$dtype()
arr$shape()
arr$chunk_shape()

# Retrieve the full array — returns an R integer matrix
data <- arr$retrieve()
dim(data)
data
```

## Subsetting (1-based, inclusive)

```{r subset}
# First two rows, first three columns
sub <- arr$retrieve(starts = c(1L, 1L), ends = c(2L, 3L))
dim(sub)
sub
```

## Array metadata

```{r meta}
arr$ndim()
arr$dimension_names()   # NULL when absent
arr$metadata()$data_type
arr$metadata()$shape
```

## Remote store (HTTP/HTTPS, S3; optional GCS/Azure)

`ZarrObjectStore` uses the Rust object-store integration underneath and
dispatches on the URL scheme. Public `https://` URLs need no credentials.
S3 is enabled by default. GCS and Azure Blob require the source-install features
shown above. For private cloud buckets, set the standard provider environment
variables before calling `open()`:

| Provider | Env vars |
|---|---|
| AWS S3 | `AWS_ACCESS_KEY_ID`, `AWS_SECRET_ACCESS_KEY`, `AWS_REGION` (+ `AWS_ENDPOINT_URL` for MinIO / custom endpoints) |
| Google Cloud | `GOOGLE_APPLICATION_CREDENTIALS` (or instance metadata) |
| Azure Blob | `AZURE_STORAGE_ACCOUNT`, `AZURE_STORAGE_ACCESS_KEY` or `AZURE_CLIENT_ID` + `AZURE_CLIENT_SECRET` |

```{r object-store}
# Public HTTPS endpoint — no credentials needed
os   <- ZarrObjectStore$open(
  "https://uk1s3.embassy.ebi.ac.uk/idr/zarr/v0.4/idr0062A/6001240.zarr"
)
oarr <- ZarrArray$open_object_store(os, "/0")
oarr$dtype()
oarr$shape()   # t, z, y, x  (4-D: 2 timepoints × 236 Z-slices × 275 × 271)

# Retrieve a small spatial patch: first time point, first Z-slice,
# first 8×8 pixels in Y/X — indices are 1-based inclusive
patch <- oarr$retrieve(c(1L, 1L, 1L, 1L), c(1L, 1L, 8L, 8L))
dim(patch)
range(patch)
```

Cloud URLs use the same API; only the scheme and credentials differ:

```r
# AWS S3 is enabled by default
Sys.setenv(
  AWS_ACCESS_KEY_ID     = "...",
  AWS_SECRET_ACCESS_KEY = "...",
  AWS_REGION            = "us-east-1"
)
s3 <- ZarrObjectStore$open("s3://my-bucket/path/to/store.zarr")

# GCS requires installing with --enable-gcp or --enable-all-cloud
Sys.setenv(GOOGLE_APPLICATION_CREDENTIALS = "/path/to/service-account.json")
gcs <- ZarrObjectStore$open("gs://my-bucket/path/to/store.zarr")

# Azure requires installing with --enable-azure or --enable-all-cloud
Sys.setenv(
  AZURE_STORAGE_ACCOUNT = "...",
  AZURE_STORAGE_ACCESS_KEY = "..."
)
az <- ZarrObjectStore$open("az://my-container/path/to/store.zarr")
```

## Group metadata

```{r group}
vcf_path <- system.file("testdata", "vcf_zarr", "v0.4", package = "Rzarrs")
grp <- ZarrGroup$open(ZarrStore$open(vcf_path), "/")
names(grp$attributes())
grp$children(FALSE)
```

## VCF Zarr reader

`ZarrVcf` reads VCF data stored in the
[VCF Zarr spec](https://github.com/sgkit-dev/vcf-zarr-spec) (v0.1–v0.4).
It accepts a local path, a URL, or an existing store handle.

```{r vcf}
vcf_path <- system.file("testdata", "vcf_zarr", "v0.4", package = "Rzarrs")
zv <- ZarrVcf$open(vcf_path)

zv$version()
zv$n_variants()
zv$n_samples()
zv$contigs()
zv$samples()
zv$variant_position()
zv$variant_contig()
zv$variant_allele()
zv$genotypes()
zv$call_genotype_phased()
```

### VCF Zarr ZIP archives

VCF Zarr data are often distributed as `.zarr.zip` archives for transfer and
small examples. `ZarrVcf$open()` accepts a local `.zarr.zip` path when the Rust
`zip` feature is enabled; `zip` is part of the default feature set.

```{r vcf-zip}
zip_path <- system.file("testdata", "vcf_zarr", "v0.4.zarr.zip", package = "Rzarrs")
zv_zip <- ZarrVcf$open(zip_path)
zv_zip$version()
zv_zip$genotypes(variants = 1:2, samples = 1:2)
```


## License

GPL (>= 3)
