---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Rzarrs

R bindings to the [`zarrs`](https://github.com/zarrs/zarrs) Rust library for
reading Zarr V3 (and compatible Zarr V2) stores — from local disk or over
HTTP/HTTPS.

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rzarrs/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rzarrs/actions/workflows/R-CMD-check.yaml)
[![Rzarrs status badge](https://sounkou-bioinfo.r-universe.dev/Rzarrs/badges/version)](https://sounkou-bioinfo.r-universe.dev/Rzarrs)
<!-- badges: end -->

## Overview

Rzarrs exposes three R6-style reference objects:

| Object | Purpose |
|---|---|
| `ZarrStore` | Local filesystem store |
| `ZarrHttpStore` | Remote store over HTTP/HTTPS |
| `ZarrArray` | A single array within any store |

Zarr dtypes are mapped to R types automatically:

| Zarr dtype | R type | Notes |
|---|---|---|
| `float32` / `float64` | `double` | NaN → `NA_real_` |
| `int8` / `int16` / `int32` | `integer` | `i32::MIN` → `NA_integer_` |
| `int64` | `double` | exact to 2^53 |
| `uint8` / `uint16` | `integer` | always fits |
| `uint32` / `uint64` | `double` | |
| `bool` | `logical` | |

SIMD-accelerated codec paths (gzip, zstd, blosc, crc32c) are selected
automatically at runtime by the underlying Rust dependency crates.

## Installation

```r
install.packages(
  "Rzarrs",
  repos = c(
    "https://sounkou-bioinfo.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

A Rust toolchain (cargo + rustc >= 1.82) is required at install time.

## Local store

```{r local}
library(Rzarrs)

# The package ships a tiny bundled fixture for illustration
path <- system.file("testdata", "int32.zarr", package = "Rzarrs")

store <- ZarrStore$open(path)
store$path()

arr <- ZarrArray$open(store, "/")
arr$dtype()
arr$shape()
arr$chunk_shape()

# Retrieve the full array — returns an R integer vector with a dim attribute
data <- arr$retrieve(NULL, NULL)
dim(data)
data
```

## Subsetting (0-based, exclusive end)

```{r subset}
# First two rows, first three columns
sub <- arr$retrieve(starts = c(0L, 0L), ends = c(2L, 3L))
dim(sub)
sub
```

## Array metadata

```{r meta}
arr$ndim()
arr$dimension_names()   # NULL when absent
jsonlite::fromJSON(arr$metadata_json())$data_type
```

## Remote store over HTTP/HTTPS

Zarr V2 OME-Zarr microscopy data from the
[Image Data Resource](https://idr.openmicroscopy.org/):

```{r http}
hs  <- ZarrHttpStore$open(
  "https://uk1s3.embassy.ebi.ac.uk/idr/zarr/v0.4/idr0062A/6001240.zarr"
)
hs$url()

img <- ZarrArray$open_http(hs, "/0")
img$dtype()
img$shape()   # t, c, z, y, x

# Fetch a small spatial patch: first time point, first channel, first Z slice
patch <- img$retrieve(
  starts = c(0L, 0L, 0L, 0L),
  ends   = c(1L, 1L, 1L, 64L)
)
dim(patch)
range(patch)
```

## License

GPL (>= 3)
