---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Rzarrs

R bindings to the [`zarrs`](https://github.com/zarrs/zarrs) Rust library for
reading Zarr V3 (and compatible Zarr V2) stores from local disk.

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rzarrs/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rzarrs/actions/workflows/R-CMD-check.yaml)
[![Rzarrs status badge](https://sounkou-bioinfo.r-universe.dev/Rzarrs/badges/version)](https://sounkou-bioinfo.r-universe.dev/Rzarrs)
<!-- badges: end -->

## Overview

Rzarrs exposes two R6-style reference objects:

| Object | Purpose |
|---|---|
| `ZarrStore` | Handle to a local filesystem store (the `.zarr` root directory) |
| `ZarrArray` | Handle to a single array within a store |

Array data is retrieved directly into R vectors with a `dim` attribute, with
automatic dtype mapping:

| Zarr dtype | R type | Notes |
|---|---|---|
| `float32` / `float64` | `double` | NaN → `NA_real_` |
| `int8` / `int16` / `int32` | `integer` | `i32::MIN` → `NA_integer_` |
| `int64` | `double` | exact to 2^53 |
| `uint8` / `uint16` | `integer` | always fits |
| `uint32` / `uint64` | `double` | |
| `bool` | `logical` | |

SIMD-accelerated codec paths (gzip, zstd, blosc, crc32c) are selected
automatically at runtime by the underlying Rust dependency crates — no
build-time feature flags required.

## Installation

```r
install.packages(
  "Rzarrs",
  repos = c(
    "https://sounkou-bioinfo.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

A Rust toolchain (cargo + rustc >= 1.82) is required at install time.

## Usage

```{r usage, eval = FALSE}
library(Rzarrs)

# Open a store
store <- ZarrStore$open("/path/to/my.zarr")
store$path()

# Open an array
arr <- ZarrArray$open(store, "/")
arr$ndim()
arr$shape()
arr$chunk_shape()
arr$dtype()

# Retrieve the full array
data <- arr$retrieve(NULL, NULL)
dim(data)

# Retrieve a subset (0-based, exclusive end)
sub <- arr$retrieve(starts = c(0L, 0L, 0L), ends = c(10L, 10L, 5L))
dim(sub)

# Inspect raw metadata
cat(arr$metadata_json())
```

## License

GPL (>= 3)
