---
title: "Rzarrs large-payload warm-read diagnostic"
output:
  github_document:
    toc: true
params:
  results_dir: !r Sys.getenv("RZARRS_LARGE_BENCH_RESULTS", unset = "")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
results_dir <- params$results_dir
stopifnot(nzchar(results_dir), dir.exists(results_dir))
results_dir <- normalizePath(results_dir, mustWork = TRUE)
```

## Question

This diagnostic increases the logical payload from 64 MiB to 1 GiB to test
whether the warm uncompressed Rzarrs--Zarrista gap is predominantly fixed
wrapper overhead. It does not attribute the remaining gap to one cause.

The fixture is a Zarr V3 `16384 x 16384` `int32` array with the bytes codec and
64 chunks of `2048 x 2048`. The logical payload is therefore 16 times larger
than the primary benchmark while the chunk count remains fixed at 64. Only the
warm bytes workload is included: gzip and cold-cache behavior answer different
questions.

Zarrista remains a Python/NumPy context point, not an R-binding comparison.
Compiler/linker parity between the installed artifacts has not been established,
so residual differences must not be described as flag-independent.

## Method

Each implementation ran in five fresh processes pinned to CPU 0 and NUMA node
0, with Blosc, Rayon, Tokio, OpenMP, and BLAS thread controls fixed at one. Each
process performed one untimed full materialization followed by three timed
open-plus-full-materialization iterations. Runtime and package startup occurred
before the timers. R used `bench::mark()` and Zarrista used
`time.perf_counter()`.

No allocation or RSS comparison is reported here. `bench` allocation tracks
cumulative R-profiled allocation rather than native or peak memory, and GNU
`time -v` covers different whole-process bookkeeping between the R and Python
drivers.

```sh
Rscript benchmarks/make_benchmark_fixtures.R \
  --out /tmp/rzarrs-large-fixtures \
  --rows 16384 --cols 16384 \
  --chunk-rows 2048 --chunk-cols 2048 --overwrite
rm -rf /tmp/rzarrs-large-fixtures/numeric-gzip.zarr

R_LIBS_USER=/tmp/rzarrs-layout-lib-final \
  benchmarks/run_rzarrs_rarr_bench.sh \
  --fixtures /tmp/rzarrs-large-fixtures \
  --out /tmp/rzarrs-large-results \
  --cpuset 0 --numa-node 0 --mode warm --reps 5 --iterations 3

benchmarks/run_zarrista_bench.sh \
  --fixtures /tmp/rzarrs-large-fixtures \
  --out /tmp/rzarrs-large-results \
  --python /tmp/zarrista-venv/bin/python \
  --zarrista-revision 92d26b65b90e9715d5c658c71b9216449f25ae64 \
  --cpuset 0 --numa-node 0 --mode warm --reps 5 --iterations 3
```

## Artifact validation

```{r read-results}
required_fields <- c(
  "implementation", "runtime", "runtime_version", "implementation_version",
  "benchmark_engine", "benchmark_engine_version", "measurement_scope",
  "startup_included", "mode", "store", "iterations_requested",
  "iterations_completed", "min_s", "median_s", "mean_s", "total_s"
)
summary_paths <- list.files(
  results_dir, pattern = "summary\\.csv$", recursive = TRUE, full.names = TRUE
)
stopifnot(length(summary_paths) == 15L)
raw_runs <- lapply(summary_paths, utils::read.csv, stringsAsFactors = FALSE)
stopifnot(all(vapply(
  raw_runs,
  function(run) nrow(run) == 1L && identical(names(run), required_fields),
  logical(1L)
)))
runs <- do.call(rbind, raw_runs)
runs$artifact <- substring(summary_paths, nchar(results_dir) + 2L)
stopifnot(
  !anyNA(runs),
  !any(vapply(runs, function(column) any(trimws(as.character(column)) == ""),
              logical(1L))),
  identical(sort(unique(runs$implementation)), c("Rarr", "Rzarrs", "Zarrista")),
  all(table(runs$implementation) == 5L),
  identical(unique(runs$mode), "warm"),
  all(runs$iterations_requested == 3L),
  all(runs$iterations_completed == 3L),
  all(tolower(as.character(runs$startup_included)) == "false")
)

manifest_paths <- file.path(dirname(summary_paths), "fixture.dcf")
stopifnot(all(file.exists(manifest_paths)))
manifests <- lapply(manifest_paths, read.dcf)
stopifnot(all(vapply(
  manifests,
  function(manifest) {
    identical(unname(manifest[1L, "shape"]), "16384x16384") &&
      identical(unname(manifest[1L, "chunk_shape"]), "2048x2048") &&
      identical(unname(manifest[1L, "chunk_count"]), "64") &&
      identical(unname(manifest[1L, "logical_bytes"]), "1073741824") &&
      identical(unname(manifest[1L, "codec"]), "bytes")
  },
  logical(1L)
)))

receipt_path <- file.path(results_dir, "fixture-receipt.txt")
stopifnot(file.exists(receipt_path))
receipt_lines <- readLines(receipt_path, warn = FALSE)
stopifnot(
  any(grepl("5a340f0358bd89df8759993cd9fd1bd04465fc27670c451217dc9291b0109127", receipt_lines, fixed = TRUE)),
  any(grepl("9e8c74b4d569351e07daf7d1787b6fff3d08f692bfbc7d2262ca0005d81af9c8", receipt_lines, fixed = TRUE))
)

environment_paths <- list.files(
  results_dir, pattern = "^environment\\.txt$", recursive = TRUE,
  full.names = TRUE
)
stopifnot(length(environment_paths) == 2L)
clean_receipt <- function(path) {
  lines <- readLines(path, warn = FALSE)
  status <- match("source_status:", lines)
  revision <- sub("^source_revision: *", "", grep(
    "^source_revision:", lines, value = TRUE
  ))
  stopifnot(
    length(revision) == 1L,
    nzchar(revision),
    !is.na(status),
    status < length(lines),
    startsWith(lines[[status + 1L]], "date_utc:")
  )
  revision
}
revisions <- vapply(environment_paths, clean_receipt, character(1L))
```

```{r receipts}
receipt_table <- data.frame(
  receipt = substring(environment_paths, nchar(results_dir) + 2L),
  clean_source_revision = revisions
)
knitr::kable(receipt_table, row.names = FALSE)
```

## Results

The per-process values below are medians of three timed iterations. Showing all
five values makes process-level variation visible without dumping raw benchmark
objects.

```{r process-results}
process_results <- runs[c("implementation", "median_s", "artifact")]
process_results$replicate <- sub(".*rep-([0-9]+)/summary\\.csv$", "\\1",
                                 process_results$artifact)
process_results <- process_results[c("implementation", "replicate", "median_s")]
process_results <- process_results[order(
  process_results$implementation, as.integer(process_results$replicate)
), ]
process_results$median_s <- signif(process_results$median_s, 7L)
knitr::kable(process_results, row.names = FALSE)
```

```{r aggregate-results}
reported <- stats::aggregate(
  runs$median_s,
  by = list(implementation = runs$implementation),
  FUN = function(value) c(
    min_s = min(value), median_s = stats::median(value), max_s = max(value)
  )
)
reported <- data.frame(
  implementation = reported$implementation,
  min_s = reported$x[, "min_s"],
  median_s = reported$x[, "median_s"],
  max_s = reported$x[, "max_s"]
)
reported$throughput_mib_s <- 1024 / reported$median_s
for (field in c("min_s", "median_s", "max_s", "throughput_mib_s")) {
  reported[[field]] <- signif(reported[[field]], 6L)
}
knitr::kable(reported, row.names = FALSE)
```

```{r ratios}
medians <- stats::setNames(reported$median_s, reported$implementation)
ratios <- data.frame(
  comparison = c("Zarrista over Rzarrs", "Rzarrs over Rarr"),
  speedup = c(
    medians[["Rzarrs"]] / medians[["Zarrista"]],
    medians[["Rarr"]] / medians[["Rzarrs"]]
  )
)
ratios$speedup <- signif(ratios$speedup, 6L)
knitr::kable(ratios, row.names = FALSE)
```

The 1 GiB medians are approximately `r sprintf("%.3f", medians[["Zarrista"]])`
seconds for Zarrista, `r sprintf("%.3f", medians[["Rzarrs"]])` seconds for
Rzarrs, and `r sprintf("%.3f", medians[["Rarr"]])` seconds for Rarr. Zarrista
is about `r sprintf("%.2f", ratios$speedup[[1L]])` times faster than Rzarrs,
while Rzarrs is about `r sprintf("%.2f", ratios$speedup[[2L]])` times faster
than Rarr. The persistence of the gap after a 16-fold payload increase argues
against a small fixed R/Python wrapper cost as the dominant explanation. It
does **not** prove that column-major transposition alone accounts for the
difference; retrieval, ownership, allocation/page-fault behavior, conversion,
and unmatched build configuration still need stage-level evidence.

## Native profile

The committed profile drivers open the array once, perform one warm
materialization, and then perform three more full materializations. `perf stat`
repeated that whole process three times; `perf record` sampled a separate run
at 499 Hz using user-space cycles. These counters include runtime startup and
array open, so they support before/after diagnosis rather than replacing the
timing table above.

```sh
perf record -F 499 -e cycles:u --call-graph fp -- \
  Rscript benchmarks/profile_large_rzarrs.R \
  --store /tmp/rzarrs-large-fixtures/numeric-uncompressed.zarr

perf record -F 499 -e cycles:u --call-graph fp -- \
  "$ZARRISTA_PYTHON" benchmarks/profile_large_zarrista.py \
  --store /tmp/rzarrs-large-fixtures/numeric-uncompressed.zarr
```

```{r profile-results}
profile_path <- "benchmark_large_payload_profile.csv"
stopifnot(file.exists(profile_path))
profile <- utils::read.csv(profile_path, stringsAsFactors = FALSE)
profile_fields <- c(
  "implementation", "mapper_variant", "source_revision",
  "native_artifact_sha256", "profile_scope", "task_clock_ms", "cycles",
  "instructions", "branches", "branch_misses", "cache_references",
  "cache_misses", "minor_faults", "major_faults", "top_resolved_symbol",
  "top_resolved_self_percent"
)
stopifnot(
  identical(names(profile), profile_fields),
  nrow(profile) == 3L,
  !anyNA(profile),
  !any(vapply(profile, function(column) {
    any(trimws(as.character(column)) == "")
  }, logical(1L))),
  identical(profile$implementation, c("Rzarrs", "Rzarrs", "Zarrista"))
)
profile_report <- profile[c(
  "implementation", "mapper_variant", "task_clock_ms", "cycles",
  "instructions", "cache_references", "cache_misses",
  "top_resolved_symbol", "top_resolved_self_percent"
)]
knitr::kable(profile_report, row.names = FALSE)
```

```{r profile-change}
before <- profile[profile$mapper_variant == "source-contiguous 32x32", ]
after <- profile[profile$mapper_variant == "destination-contiguous 32x32", ]
stopifnot(nrow(before) == 1L, nrow(after) == 1L)
fields <- c(
  "task_clock_ms", "cycles", "instructions", "branches", "branch_misses",
  "cache_references", "cache_misses", "minor_faults"
)
profile_change <- data.frame(
  metric = fields,
  before = as.numeric(before[1L, fields]),
  after = as.numeric(after[1L, fields])
)
profile_change$reduction_percent <- signif(
  100 * (profile_change$before - profile_change$after) /
    profile_change$before,
  5L
)
knitr::kable(profile_change, row.names = FALSE)
```

Changing only the inner-loop orientation reduced task-clock by about 27%,
cycles by about 50%, and observed cache misses by about 41%, while instructions
and minor faults were essentially unchanged. This is direct evidence that
write locality—not fewer conversion instructions or delayed finalization—caused
the improvement. The mapper's leading resolved self-cycle share fell from
71.33% to 42.53%. It is still the largest resolved Rzarrs hotspot; Zarrista's
leading resolved symbol is `memmove` at 60.81%, and no explicit `to_numpy`
hotspot appears in the sampled output. Symbol percentages exclude a substantial
unresolved call-chain share and must not be treated as exact wall-time stages.
