---
title: "Rzarrs versus Rarr: numeric Zarr read benchmark"
output:
  github_document:
    toc: true
params:
  results_dir: !r Sys.getenv("RZARRS_BENCH_RESULTS", unset = "")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
results_dir <- params$results_dir
has_results <- nzchar(results_dir) && dir.exists(results_dir)
```

## Scope

This report is the decision baseline before extending Rzarrs with deferred
materialization, ALTREP, or an asynchronous native API. It compares the common,
validated surface only: a complete local Zarr V3 `int32` array read into R.

The fixture generator makes two equivalent workloads:

- `numeric-uncompressed.zarr` uses only the bytes codec;
- `numeric-gzip.zarr` adds gzip level 1 after bytes.

The runner verifies bitwise-equivalent dimensions and values from Rzarrs and
Rarr before measuring either implementation. Zarr-VCF, ZIP stores, strings,
partial reads, transpose, and remote object stores are outside this baseline.
They require their own interoperable fixture and equality contract.

Zarrista is included as a separate contextual point using the same fixtures,
CPU/NUMA placement, and thread limits. It materializes a Python/NumPy array,
not an R array, so it must never be read as an R-binding comparison or folded
into the Rzarrs-versus-Rarr speedup.

## Method

The driver uses `bench::mark()` for per-process elapsed-time and allocation
measurements. Package loading occurs before `bench::mark()`; Zarrista likewise
imports its Python API before `time.perf_counter()` starts. Thus `median_s`
measures a meaningful request—open the store/array and fully materialize its
payload—not R/Python process or library startup. GNU `time -v` deliberately
remains process-inclusive, but is used only for peak RSS and CPU diagnostics,
not the elapsed-time denominator. Each implementation runs in a separate,
fresh process. The shell runner binds that process to one physical CPU and its
NUMA memory node with `taskset` and `numactl`, records GNU `time -v`, and
alternates tool order by replicate.

The baseline is deliberately single-threaded. Neither implementation has a
matched documented read-thread knob, so a multi-core taskset would not
establish equivalent thread budgets. The runner additionally pins
`BLOSC_NTHREADS`, `RAYON_NUM_THREADS`, `TOKIO_WORKER_THREADS`, and OpenMP/BLAS
thread variables to one; their exact values are recorded in `environment.txt`.
The Zarrista runner also sets `RAYON_NUM_THREADS=1`, and records its exact
upstream Git revision and Python environment. Warm and cold cache data are
distinct workloads and must never be pooled.

```sh
Rscript tools/make_benchmark_fixtures.R \
  --out "$HOME/.cache/Rzarrs/benchmarks/fixtures"

tools/run_rzarrs_rarr_bench.sh \
  --fixtures "$HOME/.cache/Rzarrs/benchmarks/fixtures" \
  --out "$HOME/.cache/Rzarrs/benchmarks/results" \
  --cpuset 2 --numa-node 0 --mode warm --reps 5 --iterations 5

sudo tools/run_rzarrs_rarr_bench.sh \
  --fixtures "$HOME/.cache/Rzarrs/benchmarks/fixtures" \
  --out "$HOME/.cache/Rzarrs/benchmarks/results" \
  --cpuset 2 --numa-node 0 --mode cold --reps 5

# ZARRISTA_PYTHON has zarrista + numpy installed from ZARRISTA_REVISION.
tools/run_zarrista_bench.sh \
  --fixtures "$HOME/.cache/Rzarrs/benchmarks/fixtures" \
  --out "$HOME/.cache/Rzarrs/benchmarks/results" \
  --python "$ZARRISTA_PYTHON" --zarrista-revision "$ZARRISTA_REVISION" \
  --cpuset 2 --numa-node 0 --mode warm --reps 5 --iterations 5

RZARRS_BENCH_RESULTS="$HOME/.cache/Rzarrs/benchmarks/results" \
  Rscript -e 'rmarkdown::render("benchmarks/benchmark_rzarrs_rarr.Rmd")'
```

## Recorded environment

Every run directory contains `bench.rds`, `summary.csv`, GNU `time -v` output,
the exact command line, R session information, and its fixture manifest. Each
cache-mode directory has `environment.txt`, including source revision/status,
CPU and NUMA topology, package versions, and run configuration. The report
below only aggregates artifacts that share this result directory.

```{r read-results}
read_time_v <- function(path) {
  if (!file.exists(path)) return(c(max_rss_mib = NA_real_, cpu_percent = NA_real_))
  lines <- readLines(path, warn = FALSE)
  rss_line <- grep("Maximum resident set size", lines, value = TRUE)
  cpu_line <- grep("Percent of CPU this job got", lines, value = TRUE)
  rss_kib <- if (length(rss_line)) {
    as.numeric(sub(".*: *", "", rss_line[[1L]]))
  } else {
    NA_real_
  }
  cpu_percent <- if (length(cpu_line)) {
    as.numeric(sub("%.*", "", sub(".*: *", "", cpu_line[[1L]])))
  } else {
    NA_real_
  }
  c(max_rss_mib = rss_kib / 1024, cpu_percent = cpu_percent)
}

read_manifest <- function(path) {
  manifest <- file.path(dirname(path), "fixture.dcf")
  if (!file.exists(manifest)) return(c(logical_bytes = NA_real_, codec = NA_character_))
  lines <- trimws(readLines(manifest, warn = FALSE))
  pairs <- regmatches(lines, regexec("^([^:=]+)[:=] *(.*)$", lines))
  if (any(lengths(pairs) != 3L)) stop("malformed fixture manifest: ", manifest)
  fields <- stats::setNames(
    vapply(pairs, `[[`, character(1L), 3L),
    vapply(pairs, `[[`, character(1L), 2L)
  )
  c(
    logical_bytes = as.numeric(fields[["logical_bytes"]]),
    codec = unname(fields[["codec"]])
  )
}

read_run <- function(path) {
  result <- utils::read.csv(path, stringsAsFactors = FALSE)
  if (!"runtime" %in% names(result)) result$runtime <- "R"
  common_fields <- c(
    "implementation", "runtime", "mode", "store", "iterations_requested",
    "iterations_completed", "min_s", "median_s", "mean_s", "total_s",
    "mem_alloc_bytes", "gc_count", "r_version", "rzarrs_version",
    "rarr_version", "bench_version"
  )
  for (field in setdiff(common_fields, names(result))) result[[field]] <- NA
  result <- result[common_fields]
  result$run_dir <- dirname(path)
  result$fixture <- basename(result$store)
  manifest <- read_manifest(path)
  result$logical_bytes <- as.numeric(manifest[["logical_bytes"]])
  result$codec <- manifest[["codec"]]
  resources <- read_time_v(file.path(result$run_dir, "time-v.txt"))
  result$max_rss_mib <- resources[["max_rss_mib"]]
  result$cpu_percent <- resources[["cpu_percent"]]
  result
}

if (has_results) {
  summary_paths <- list.files(
    results_dir, pattern = "summary\\.csv$", recursive = TRUE, full.names = TRUE
  )
  if (!length(summary_paths)) stop("no summary.csv files under: ", results_dir)
  runs <- do.call(rbind, lapply(summary_paths, read_run))
  runs$throughput_mib_s <- runs$logical_bytes / runs$median_s / 1024^2
  runs
} else {
  data.frame()
}
r_runs <- if (nrow(runs)) subset(runs, runtime == "R") else runs
zarrista_runs <- if (nrow(runs)) subset(runs, implementation == "Zarrista") else runs
```

```{r environment, results='asis'}
if (has_results) {
  environment_paths <- list.files(
    results_dir, pattern = "^environment\\.txt$", recursive = TRUE, full.names = TRUE
  )
  if (!length(environment_paths)) stop("no environment.txt files under: ", results_dir)
  for (environment_path in environment_paths) {
    cat("### ", basename(dirname(environment_path)), " environment\n\n", sep = "")
    cat("```text\n")
    cat(readLines(environment_path, warn = FALSE), sep = "\n")
    cat("\n```\n\n")
  }
} else {
  cat("Render with `RZARRS_BENCH_RESULTS` set to a runner output directory.\n")
}
```

## Results

`bench` repeats inside a process. The table therefore reports the median of
each process's `bench` median across process-level replicates. `max_rss_mib`
and `cpu_percent` are likewise process-level medians from GNU `time -v`.

```{r results-table}
if (nrow(r_runs)) {
  reported <- stats::aggregate(
    r_runs[c("median_s", "mem_alloc_bytes", "throughput_mib_s", "max_rss_mib", "cpu_percent")],
    by = list(
      mode = r_runs$mode,
      fixture = r_runs$fixture,
      codec = r_runs$codec,
      implementation = r_runs$implementation
    ),
    FUN = stats::median,
    na.rm = TRUE
  )
  names(reported)[names(reported) == "Group.1"] <- "mode"
  names(reported)[names(reported) == "Group.2"] <- "fixture"
  names(reported)[names(reported) == "Group.3"] <- "codec"
  names(reported)[names(reported) == "Group.4"] <- "implementation"
  reported$median_s <- signif(reported$median_s, 5)
  reported$mem_alloc_mib <- signif(reported$mem_alloc_bytes / 1024^2, 5)
  reported$throughput_mib_s <- signif(reported$throughput_mib_s, 5)
  reported$max_rss_mib <- signif(reported$max_rss_mib, 5)
  reported$cpu_percent <- signif(reported$cpu_percent, 5)
  reported$mem_alloc_bytes <- NULL
  knitr::kable(reported, row.names = FALSE)
} else {
  cat("No benchmark artifacts supplied.\n")
}
```

A speedup is only meaningful within one `mode`, fixture, codec, CPU binding,
and environment. For each matched pair below, a value above 1 means Rzarrs had
the lower process-level `bench` median.

```{r speedup}
if (nrow(r_runs)) {
  per_run <- stats::aggregate(
    r_runs["median_s"],
    by = list(mode = r_runs$mode, fixture = r_runs$fixture, codec = r_runs$codec,
              implementation = r_runs$implementation),
    FUN = stats::median
  )
  wide <- reshape(
    per_run,
    idvar = c("mode", "fixture", "codec"),
    timevar = "implementation",
    direction = "wide"
  )
  if (all(c("median_s.Rzarrs", "median_s.Rarr") %in% names(wide))) {
    wide$rzarrs_speedup_over_rarr <- wide$median_s.Rarr / wide$median_s.Rzarrs
    wide$rzarrs_speedup_over_rarr <- signif(wide$rzarrs_speedup_over_rarr, 5)
    knitr::kable(wide, row.names = FALSE)
  } else {
    cat("Both implementations are required for a speedup calculation.\n")
  }
}
```

## Zarrista context baseline

Zarrista runs the same local V3 fixtures with the same CPU/NUMA and native
thread controls, but its result is a NumPy array. The table intentionally omits
R allocation measurements and does not calculate an Rzarrs/Rarr-style speedup.
It is a comparison point for a Rust-native Python reader, not a prospective R
binding measurement.

```{r zarrista-results}
if (nrow(zarrista_runs)) {
  zarrista_report <- stats::aggregate(
    zarrista_runs[c("median_s", "throughput_mib_s", "max_rss_mib", "cpu_percent")],
    by = list(
      mode = zarrista_runs$mode,
      fixture = zarrista_runs$fixture,
      codec = zarrista_runs$codec
    ),
    FUN = stats::median,
    na.rm = TRUE
  )
  for (field in c("median_s", "throughput_mib_s", "max_rss_mib", "cpu_percent")) {
    zarrista_report[[field]] <- signif(as.numeric(zarrista_report[[field]]), 5)
  }
  knitr::kable(zarrista_report, row.names = FALSE)
} else {
  cat("No Zarrista artifacts supplied.\n")
}
```

## Visual comparisons

These plots include all three readers. Zarrista remains a Python/NumPy context
point; visual proximity is not an R-binding claim. The timed operation is the
loaded-runtime, open-and-materialize request recorded in `measurement_scope`.

```{r reader-plots, fig.width=11, fig.height=7}
if (nrow(r_runs) && nrow(zarrista_runs)) {
  plot_runs <- rbind(
    r_runs[c("implementation", "mode", "codec", "median_s", "throughput_mib_s", "max_rss_mib")],
    zarrista_runs[c("implementation", "mode", "codec", "median_s", "throughput_mib_s", "max_rss_mib")]
  )
  plot_runs$workload <- paste(plot_runs$mode, plot_runs$codec, sep = " / ")
  summary_runs <- stats::aggregate(
    plot_runs[c("median_s", "throughput_mib_s", "max_rss_mib")],
    by = list(implementation = plot_runs$implementation, workload = plot_runs$workload),
    FUN = stats::median,
    na.rm = TRUE
  )
  implementations <- c("Rarr", "Rzarrs", "Zarrista")
  workloads <- c("cold / bytes", "warm / bytes", "cold / gzip(level=1)", "warm / gzip(level=1)")
  colours <- c(Rarr = "#0072B2", Rzarrs = "#D55E00", Zarrista = "#009E73")

  draw_metric <- function(metric, label) {
    old <- par(mfrow = c(2, 2), mar = c(4, 8, 3, 2))
    on.exit(par(old), add = TRUE)
    for (workload in workloads) {
      values <- summary_runs[summary_runs$workload == workload, c("implementation", metric)]
      values <- values[match(implementations, values$implementation), ]
      values <- values[!is.na(values[[metric]]), ]
      x <- values[[metric]]
      pad <- c(min(x) / 1.8, max(x) * 1.8)
      plot(x, seq_along(x), log = "x", xlim = pad, yaxt = "n", pch = 19,
           col = colours[values$implementation], xlab = label, ylab = "", main = workload)
      axis(2, at = seq_along(x), labels = values$implementation, las = 1)
      text(x, seq_along(x), labels = format(signif(x, 4), trim = TRUE), pos = 4, cex = 0.8)
      grid(col = "grey90")
    }
  }

  draw_metric("median_s", "Median seconds (lower is better; log scale)")
  draw_metric("throughput_mib_s", "MiB/s (higher is better; log scale)")
  draw_metric("max_rss_mib", "Maximum RSS MiB (process diagnostic; log scale)")
} else {
  cat("Rarr, Rzarrs, and Zarrista artifacts are all required for plots.\n")
}
```

## Network-bounded S3 matrix

The local-store table above is not a network result. The remote version is a
separate S3-compatible workload: stage the immutable fixture once into a local
MinIO bucket, run each client in a dedicated network namespace, and apply the
same `tc netem` rate and latency in both directions on its veth pair. Record
MinIO revision, bucket/prefix, object count/bytes, endpoint, latency, rate,
page-cache mode, and client thread limits. Do not use an unbounded public
endpoint or fold local and remote results together.

It is gated on a successful equivalent full-array remote probe for all three
clients: Rarr through its explicit `s3_client`, Rzarrs through
`ZarrObjectStore`, and Zarrista through an Obstore S3 store. Until that probe
is executable and equality-checked, the report must say that an all-three
network comparison is unavailable rather than presenting a two-reader result
as equivalent.

## Compilation provenance and controlled rebuild gate

The environment blocks record toolchains, binary hashes, and build-affecting
environment variables. They identify what ran, but compiler identity alone
does **not** prove flag parity—especially for an already-installed Rarr binary.
Do not attribute the observed gap to library design until the controlled rebuild
campaign below reproduces it.

1. Build Rarr and Rzarrs from source in a clean R environment with no user or
   site Makevars overrides; save every compile and link command.
2. Pin and record the Rust toolchain, Cargo release profile, C/C++ compiler,
   C/C++ flags, linker flags, enabled features, and binary SHA-256 hashes.
3. Re-run this exact fixture, CPU/NUMA/thread/cache matrix on those artifacts.
4. Report both the installed-artifact and controlled-rebuild tables; only a
   stable gap across them rules out unknown compilation flags as the cause.

## Interpretation limits

This report attributes no cost by itself. The difference between bytes-only and
gzip fixtures estimates the combined codec contribution for this workload, not
compression time in isolation. Profiling comes next only if the measured result
makes an extension worthwhile; it must distinguish I/O, decompression, data
layout conversion, and R object materialization.
