---
title: "Rzarrs large-payload warm-read diagnostic"
output:
  github_document:
    toc: true
params:
  results_dir: !r Sys.getenv("RZARRS_LARGE_BENCH_RESULTS", unset = "")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
results_dir <- params$results_dir
stopifnot(nzchar(results_dir), dir.exists(results_dir))
results_dir <- normalizePath(results_dir, mustWork = TRUE)
```

## Question

This diagnostic increases the logical payload from 64 MiB to 1 GiB to test
whether the warm uncompressed Rzarrs--Zarrista gap is predominantly fixed
wrapper overhead. It does not attribute the remaining gap to one cause.

The fixture is a Zarr V3 `16384 x 16384` `int32` array with the bytes codec and
64 chunks of `2048 x 2048`. The logical payload is therefore 16 times larger
than the primary benchmark while the chunk count remains fixed at 64. Only the
warm bytes workload is included: gzip and cold-cache behavior answer different
questions.

Zarrista remains a Python/NumPy context point, not an R-binding comparison.
Compiler/linker parity between the installed artifacts has not been established,
so residual differences must not be described as flag-independent.

## Method

Each implementation ran in five fresh processes pinned to CPU 0 and NUMA node
0, with Blosc, Rayon, Tokio, OpenMP, and BLAS thread controls fixed at one. Each
process performed one untimed full materialization followed by three timed
open-plus-full-materialization iterations. Runtime and package startup occurred
before the timers. R used `bench::mark()` and Zarrista used
`time.perf_counter()`.

No allocation or RSS comparison is reported here. `bench` allocation tracks
cumulative R-profiled allocation rather than native or peak memory, and GNU
`time -v` covers different whole-process bookkeeping between the R and Python
drivers.

```sh
Rscript benchmarks/make_benchmark_fixtures.R \
  --out /tmp/rzarrs-large-fixtures \
  --rows 16384 --cols 16384 \
  --chunk-rows 2048 --chunk-cols 2048 --overwrite
rm -rf /tmp/rzarrs-large-fixtures/numeric-gzip.zarr

R_LIBS_USER=/tmp/rzarrs-layout-lib-final \
  benchmarks/run_rzarrs_rarr_bench.sh \
  --fixtures /tmp/rzarrs-large-fixtures \
  --out /tmp/rzarrs-large-results \
  --cpuset 0 --numa-node 0 --mode warm --reps 5 --iterations 3

benchmarks/run_zarrista_bench.sh \
  --fixtures /tmp/rzarrs-large-fixtures \
  --out /tmp/rzarrs-large-results \
  --python /tmp/zarrista-venv/bin/python \
  --zarrista-revision 92d26b65b90e9715d5c658c71b9216449f25ae64 \
  --cpuset 0 --numa-node 0 --mode warm --reps 5 --iterations 3
```

## Artifact validation

```{r read-results}
required_fields <- c(
  "implementation", "runtime", "runtime_version", "implementation_version",
  "benchmark_engine", "benchmark_engine_version", "measurement_scope",
  "startup_included", "mode", "store", "iterations_requested",
  "iterations_completed", "min_s", "median_s", "mean_s", "total_s"
)
summary_paths <- list.files(
  results_dir, pattern = "summary\\.csv$", recursive = TRUE, full.names = TRUE
)
stopifnot(length(summary_paths) == 15L)
raw_runs <- lapply(summary_paths, utils::read.csv, stringsAsFactors = FALSE)
stopifnot(all(vapply(
  raw_runs,
  function(run) nrow(run) == 1L && identical(names(run), required_fields),
  logical(1L)
)))
runs <- do.call(rbind, raw_runs)
runs$artifact <- substring(summary_paths, nchar(results_dir) + 2L)
stopifnot(
  !anyNA(runs),
  !any(vapply(runs, function(column) any(trimws(as.character(column)) == ""),
              logical(1L))),
  identical(sort(unique(runs$implementation)), c("Rarr", "Rzarrs", "Zarrista")),
  all(table(runs$implementation) == 5L),
  identical(unique(runs$mode), "warm"),
  all(runs$iterations_requested == 3L),
  all(runs$iterations_completed == 3L),
  all(tolower(as.character(runs$startup_included)) == "false")
)

manifest_paths <- file.path(dirname(summary_paths), "fixture.dcf")
stopifnot(all(file.exists(manifest_paths)))
manifests <- lapply(manifest_paths, read.dcf)
stopifnot(all(vapply(
  manifests,
  function(manifest) {
    identical(unname(manifest[1L, "shape"]), "16384x16384") &&
      identical(unname(manifest[1L, "chunk_shape"]), "2048x2048") &&
      identical(unname(manifest[1L, "chunk_count"]), "64") &&
      identical(unname(manifest[1L, "logical_bytes"]), "1073741824") &&
      identical(unname(manifest[1L, "codec"]), "bytes")
  },
  logical(1L)
)))

receipt_path <- file.path(results_dir, "fixture-receipt.txt")
stopifnot(file.exists(receipt_path))
receipt_lines <- readLines(receipt_path, warn = FALSE)
stopifnot(
  any(grepl("5a340f0358bd89df8759993cd9fd1bd04465fc27670c451217dc9291b0109127", receipt_lines, fixed = TRUE)),
  any(grepl("9e8c74b4d569351e07daf7d1787b6fff3d08f692bfbc7d2262ca0005d81af9c8", receipt_lines, fixed = TRUE))
)

environment_paths <- list.files(
  results_dir, pattern = "^environment\\.txt$", recursive = TRUE,
  full.names = TRUE
)
stopifnot(length(environment_paths) == 2L)
clean_receipt <- function(path) {
  lines <- readLines(path, warn = FALSE)
  status <- match("source_status:", lines)
  revision <- sub("^source_revision: *", "", grep(
    "^source_revision:", lines, value = TRUE
  ))
  stopifnot(
    length(revision) == 1L,
    nzchar(revision),
    !is.na(status),
    status < length(lines),
    startsWith(lines[[status + 1L]], "date_utc:")
  )
  revision
}
revisions <- vapply(environment_paths, clean_receipt, character(1L))
```

```{r receipts}
receipt_table <- data.frame(
  receipt = substring(environment_paths, nchar(results_dir) + 2L),
  clean_source_revision = revisions
)
knitr::kable(receipt_table, row.names = FALSE)
```

## Results

The per-process values below are medians of three timed iterations. Showing all
five values makes process-level variation visible without dumping raw benchmark
objects.

```{r process-results}
process_results <- runs[c("implementation", "median_s", "artifact")]
process_results$replicate <- sub(".*rep-([0-9]+)/summary\\.csv$", "\\1",
                                 process_results$artifact)
process_results <- process_results[c("implementation", "replicate", "median_s")]
process_results <- process_results[order(
  process_results$implementation, as.integer(process_results$replicate)
), ]
process_results$median_s <- signif(process_results$median_s, 7L)
knitr::kable(process_results, row.names = FALSE)
```

```{r aggregate-results}
reported <- stats::aggregate(
  runs$median_s,
  by = list(implementation = runs$implementation),
  FUN = function(value) c(
    min_s = min(value), median_s = stats::median(value), max_s = max(value)
  )
)
reported <- data.frame(
  implementation = reported$implementation,
  min_s = reported$x[, "min_s"],
  median_s = reported$x[, "median_s"],
  max_s = reported$x[, "max_s"]
)
reported$throughput_mib_s <- 1024 / reported$median_s
for (field in c("min_s", "median_s", "max_s", "throughput_mib_s")) {
  reported[[field]] <- signif(reported[[field]], 6L)
}
knitr::kable(reported, row.names = FALSE)
```

```{r ratios}
medians <- stats::setNames(reported$median_s, reported$implementation)
ratios <- data.frame(
  comparison = c("Zarrista over Rzarrs", "Rzarrs over Rarr"),
  speedup = c(
    medians[["Rzarrs"]] / medians[["Zarrista"]],
    medians[["Rarr"]] / medians[["Rzarrs"]]
  )
)
ratios$speedup <- signif(ratios$speedup, 6L)
knitr::kable(ratios, row.names = FALSE)
```

The 1 GiB medians are approximately 0.432 seconds for Zarrista, 1.832 seconds
for Rzarrs, and 3.614 seconds for Rarr. Zarrista remains about 4.25 times faster
than Rzarrs, while Rzarrs is about 1.97 times faster than Rarr. The persistence
of the gap after a 16-fold payload increase argues against a small fixed
R/Python wrapper cost as the dominant explanation. It does **not** prove that
column-major transposition alone accounts for the difference; retrieval,
ownership, allocation/page-fault behavior, conversion, and unmatched build
configuration still need stage-level evidence.
