---
title: "Rzarrs versus Rarr: numeric Zarr read benchmark"
output:
  github_document:
    toc: true
params:
  results_dir: !r Sys.getenv("RZARRS_BENCH_RESULTS", unset = "")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
results_dir <- params$results_dir
has_results <- nzchar(results_dir) && dir.exists(results_dir)
```

## Scope

This report is the decision baseline before extending Rzarrs with deferred
materialization, ALTREP, or an asynchronous native API. It compares the common,
validated surface only: a complete local Zarr V3 `int32` array read into R.

The fixture generator makes two equivalent workloads:

- `numeric-uncompressed.zarr` uses only the bytes codec;
- `numeric-gzip.zarr` adds gzip level 1 after bytes.

The runner verifies bitwise-equivalent dimensions and values from Rzarrs and
Rarr before measuring either implementation. Zarr-VCF, ZIP stores, strings,
partial reads, transpose, and remote object stores are outside this baseline.
They require their own interoperable fixture and equality contract.

## Method

The driver uses `bench::mark()` for per-process elapsed-time and allocation
measurements. Each implementation runs in a separate, fresh R process. The
shell runner binds that process to one physical CPU and its NUMA memory node
with `taskset` and `numactl`, records GNU `time -v`, and alternates tool order
by replicate.

The baseline is deliberately single-threaded. Neither implementation has a
matched documented read-thread knob, so a multi-core taskset would not
establish equivalent thread budgets. The runner additionally pins
`BLOSC_NTHREADS`, `RAYON_NUM_THREADS`, `TOKIO_WORKER_THREADS`, and OpenMP/BLAS
thread variables to one; their exact values are recorded in `environment.txt`.
Warm and cold cache data are distinct workloads and must never be pooled.

```sh
Rscript tools/make_benchmark_fixtures.R \
  --out "$HOME/.cache/Rzarrs/benchmarks/fixtures"

tools/run_rzarrs_rarr_bench.sh \
  --fixtures "$HOME/.cache/Rzarrs/benchmarks/fixtures" \
  --out "$HOME/.cache/Rzarrs/benchmarks/results" \
  --cpuset 2 --numa-node 0 --mode warm --reps 5 --iterations 5

sudo tools/run_rzarrs_rarr_bench.sh \
  --fixtures "$HOME/.cache/Rzarrs/benchmarks/fixtures" \
  --out "$HOME/.cache/Rzarrs/benchmarks/results" \
  --cpuset 2 --numa-node 0 --mode cold --reps 5

RZARRS_BENCH_RESULTS="$HOME/.cache/Rzarrs/benchmarks/results" \
  Rscript -e 'rmarkdown::render("benchmarks/benchmark_rzarrs_rarr.Rmd")'
```

## Recorded environment

Every run directory contains `bench.rds`, `summary.csv`, GNU `time -v` output,
the exact command line, R session information, and its fixture manifest. Each
cache-mode directory has `environment.txt`, including source revision/status,
CPU and NUMA topology, package versions, and run configuration. The report
below only aggregates artifacts that share this result directory.

```{r read-results}
read_time_v <- function(path) {
  if (!file.exists(path)) return(c(max_rss_mib = NA_real_, cpu_percent = NA_real_))
  lines <- readLines(path, warn = FALSE)
  rss_line <- grep("Maximum resident set size", lines, value = TRUE)
  cpu_line <- grep("Percent of CPU this job got", lines, value = TRUE)
  rss_kib <- if (length(rss_line)) {
    as.numeric(sub(".*: *", "", rss_line[[1L]]))
  } else {
    NA_real_
  }
  cpu_percent <- if (length(cpu_line)) {
    as.numeric(sub("%.*", "", sub(".*: *", "", cpu_line[[1L]])))
  } else {
    NA_real_
  }
  c(max_rss_mib = rss_kib / 1024, cpu_percent = cpu_percent)
}

read_manifest <- function(path) {
  manifest <- file.path(dirname(path), "fixture.dcf")
  if (!file.exists(manifest)) return(c(logical_bytes = NA_real_, codec = NA_character_))
  lines <- trimws(readLines(manifest, warn = FALSE))
  pairs <- regmatches(lines, regexec("^([^:=]+)[:=] *(.*)$", lines))
  if (any(lengths(pairs) != 3L)) stop("malformed fixture manifest: ", manifest)
  fields <- stats::setNames(
    vapply(pairs, `[[`, character(1L), 3L),
    vapply(pairs, `[[`, character(1L), 2L)
  )
  c(
    logical_bytes = as.numeric(fields[["logical_bytes"]]),
    codec = unname(fields[["codec"]])
  )
}

read_run <- function(path) {
  result <- utils::read.csv(path, stringsAsFactors = FALSE)
  result$run_dir <- dirname(path)
  result$fixture <- basename(result$store)
  manifest <- read_manifest(path)
  result$logical_bytes <- as.numeric(manifest[["logical_bytes"]])
  result$codec <- manifest[["codec"]]
  resources <- read_time_v(file.path(result$run_dir, "time-v.txt"))
  result$max_rss_mib <- resources[["max_rss_mib"]]
  result$cpu_percent <- resources[["cpu_percent"]]
  result
}

if (has_results) {
  summary_paths <- list.files(
    results_dir, pattern = "summary\\.csv$", recursive = TRUE, full.names = TRUE
  )
  if (!length(summary_paths)) stop("no summary.csv files under: ", results_dir)
  runs <- do.call(rbind, lapply(summary_paths, read_run))
  runs$throughput_mib_s <- runs$logical_bytes / runs$median_s / 1024^2
  runs
} else {
  data.frame()
}
```

```{r environment, results='asis'}
if (has_results) {
  environment_paths <- list.files(
    results_dir, pattern = "^environment\\.txt$", recursive = TRUE, full.names = TRUE
  )
  if (!length(environment_paths)) stop("no environment.txt files under: ", results_dir)
  for (environment_path in environment_paths) {
    cat("### ", basename(dirname(environment_path)), " environment\n\n", sep = "")
    cat("```text\n")
    cat(readLines(environment_path, warn = FALSE), sep = "\n")
    cat("\n```\n\n")
  }
} else {
  cat("Render with `RZARRS_BENCH_RESULTS` set to a runner output directory.\n")
}
```

## Results

`bench` repeats inside a process. The table therefore reports the median of
each process's `bench` median across process-level replicates. `max_rss_mib`
and `cpu_percent` are likewise process-level medians from GNU `time -v`.

```{r results-table}
if (nrow(runs)) {
  reported <- stats::aggregate(
    runs[c("median_s", "mem_alloc_bytes", "throughput_mib_s", "max_rss_mib", "cpu_percent")],
    by = list(
      mode = runs$mode,
      fixture = runs$fixture,
      codec = runs$codec,
      implementation = runs$implementation
    ),
    FUN = stats::median,
    na.rm = TRUE
  )
  names(reported)[names(reported) == "Group.1"] <- "mode"
  names(reported)[names(reported) == "Group.2"] <- "fixture"
  names(reported)[names(reported) == "Group.3"] <- "codec"
  names(reported)[names(reported) == "Group.4"] <- "implementation"
  reported$median_s <- signif(reported$median_s, 5)
  reported$mem_alloc_mib <- signif(reported$mem_alloc_bytes / 1024^2, 5)
  reported$throughput_mib_s <- signif(reported$throughput_mib_s, 5)
  reported$max_rss_mib <- signif(reported$max_rss_mib, 5)
  reported$cpu_percent <- signif(reported$cpu_percent, 5)
  reported$mem_alloc_bytes <- NULL
  knitr::kable(reported, row.names = FALSE)
} else {
  cat("No benchmark artifacts supplied.\n")
}
```

A speedup is only meaningful within one `mode`, fixture, codec, CPU binding,
and environment. For each matched pair below, a value above 1 means Rzarrs had
the lower process-level `bench` median.

```{r speedup}
if (nrow(runs)) {
  per_run <- stats::aggregate(
    runs["median_s"],
    by = list(mode = runs$mode, fixture = runs$fixture, codec = runs$codec,
              implementation = runs$implementation),
    FUN = stats::median
  )
  wide <- reshape(
    per_run,
    idvar = c("mode", "fixture", "codec"),
    timevar = "implementation",
    direction = "wide"
  )
  if (all(c("median_s.Rzarrs", "median_s.Rarr") %in% names(wide))) {
    wide$rzarrs_speedup_over_rarr <- wide$median_s.Rarr / wide$median_s.Rzarrs
    wide$rzarrs_speedup_over_rarr <- signif(wide$rzarrs_speedup_over_rarr, 5)
    knitr::kable(wide, row.names = FALSE)
  } else {
    cat("Both implementations are required for a speedup calculation.\n")
  }
}
```

## Interpretation limits

This report attributes no cost by itself. The difference between bytes-only and
gzip fixtures estimates the combined codec contribution for this workload, not
compression time in isolation. Profiling comes next only if the measured result
makes an extension worthwhile; it must distinguish I/O, decompression, data
layout conversion, and R object materialization.
